Supermartingales and Termination Analysis of Statistical PCF

### 2021-10-01 LICS21/antitone.tex
Replace "theorem" environment with "therm", to get numbering back.

\begin{restatable}{theorem}{ACStrict}
\label{thm:a-c strict}
Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale w.r.t.~stopping time $T$. 
Then $T < \infty$ a.s.
\end{restatable}

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed PPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
Theorem \ref{thm:rankable implies termination},
take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same r.v.~$T_n, X_n, Y_n, T_M^{\tY}$. 
Thanks to Theorem \ref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$. 
Thus, by Theorem \ref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

\begin{restatable}[Sparse function]{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

@@

The following theorem does not prove completeness, but is suggestive in that direction:
%\begin{restatable}{theorem}{antitoneSeemsComplete}
\begin{theorem}
\label{thm:towards completeness}
For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{theorem}
%\end{restatable}

>>>>

\begin{restatable}{therm}{ACStrict}
\label{thm:a-c strict}
Let $(Y_n)_{n \geq 0}$ be an antitone strict supermartingale w.r.t.~stopping time $T$. 
Then $T < \infty$ a.s.
\end{restatable}

@@

\begin{therm}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed PPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{therm}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
Theorem \ref{thm:rankable implies termination},
take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same r.v.~$T_n, X_n, Y_n, T_M^{\tY}$. 
Thanks to Theorem \ref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$. 
Thus, by Theorem \ref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

\begin{restatable}[Sparse function]{therm}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

@@

The following theorem does not prove completeness, but is suggestive in that direction:
%\begin{restatable}{therm}{antitoneSeemsComplete}
\begin{therm}
\label{thm:towards completeness}
For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{therm}
%\end{restatable}

### 2021-10-01 LICS21/confluent.tex
Replace "theorem" environment with "therm", to get numbering back.

\begin{restatable}{theorem}{AstEquivalence} \label{thm:AstEquivalence}
A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
\end{restatable}

\begin{restatable}{theorem}{CbvIsTerminatingest} \label{thm:CbvIsTerminatingest}
If $M$ terminates with some reduction strategy $f$ and trace $s$, it terminates with $\cbv$ and $s$.
\end{restatable}

@@

All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too. The proofs of Theorem \ref{thm:rankable implies termination,thm:partial implies rankable,thm:antitone rankable implies termination,thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies. They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy). For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x. O]$, and $r(N)$ being undefined for $N$ being a value.

@@

\begin{theorem}[Confluent ranking] \label{thm:confluent ranking}
If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function w.r.t.~a reduction strategy $r$, then $M$ is AST w.r.t.~$r$, and AST.
\end{theorem}

@@

\begin{restatable}{theorem}{confluentNoStronger}
\label{thm:confluentNoStronger}
For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $\mathit{Rch}_r(M)$ is AST, then every term in $\mathit{Rch}(M)$ is AST.
\end{restatable}

>>>>

\begin{restatable}{therm}{AstEquivalence} \label{thm:AstEquivalence}
A closed term $M$ is AST with respect to $\cbv$ iff it is AST.
\end{restatable}

\begin{restatable}{therm}{CbvIsTerminatingest} \label{thm:CbvIsTerminatingest}
If $M$ terminates with some reduction strategy $f$ and trace $s$, it terminates with $\cbv$ and $s$.
\end{restatable}

@@

All of the theorems on the termination of rankable terms therefore extend to other reduction strategies too. The proofs of Theorems \ref{thm:rankable implies termination}, \ref{thm:partial implies rankable}, \ref{thm:antitone rankable implies termination} and \ref{thm:antitone partial implies rankable} are all sufficiently generic with respect to what the reduction relation actually is that they can be directly applied to other reduction strategies. They only require that the number of reductions that can occur without any of them being a $\tY$-reduction is bounded for any starting term (which is true, because Theorem \ref{thm:de groote} applies equally to any reduction strategy). For a reduction strategy $r$ and term $M$, just substitute $N[r(N)]$ being a $\tY$-redex for $N$ being of the form $E[\tY \lambda x. O]$, and $r(N)$ being undefined for $N$ being a value.

@@

\begin{therm}[Confluent ranking] \label{thm:confluent ranking}
If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function w.r.t.~a reduction strategy $r$, then $M$ is AST w.r.t.~$r$, and AST.
\end{therm}

@@

\begin{restatable}{therm}{confluentNoStronger}
\label{thm:confluentNoStronger}
For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $\mathit{Rch}_r(M)$ is AST, then every term in $\mathit{Rch}(M)$ is AST.
\end{restatable}

### 2021-10-01 LICS21/intro.tex
Replace "theorem" environment with "therm", to get numbering back.

By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples \ref{ex:non-affine recursion easy,ex:non-affine continuous}) and recursions that define higher-order functions (Example \ref{ex:higher-order recursion}).

>>>>

By combining these methods we can prove AST of a variety of PPCF programs that are beyond the reach of methods in the literature, including non-affine recursion\footnote{\label{fnote:non-affine} Non-affine recursive programs are recursive programs that can, during the evaluation of the recursive body, make multiple recursive calls (of a first-order function) from distinct call sites.} (Examples \ref{ex:non-affine recursion easy} and \ref{ex:non-affine continuous}) and recursions that define higher-order functions (Example \ref{ex:higher-order recursion}).

### 2021-10-01 LICS21/ranking.tex
Replace "theorem" environment with "therm", to get numbering back.

\begin{restatable}{theorem}{ThmMinimal} \label{thm:minimal}
Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
\[
f(N) := \mathbb E [\text{number of }\tY\text{-reduction steps from }N\text{ to a value}]
\] 
if it exists, is the least of all possible ranking functions of $M$.
\end{restatable}

@@

\begin{restatable}[Sparse function]{theorem}{PartialImpliesRankable} \label{thm:partial implies rankable}
Every sparse ranking function is a restriction of a ranking function.
\end{restatable}

>>>>

\begin{restatable}{therm}{ThmMinimal} \label{thm:minimal}
Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
\[
f(N) := \mathbb E [\text{number of }\tY\text{-reduction steps from }N\text{ to a value}]
\] 
if it exists, is the least of all possible ranking functions of $M$.
\end{restatable}

@@

\begin{restatable}[Sparse function]{therm}{PartialImpliesRankable} \label{thm:partial implies rankable}
Every sparse ranking function is a restriction of a ranking function.
\end{restatable}

### 2021-10-01 LICS21/related.tex
Replace "theorem" environment with "therm", to get numbering back.

\cite{DBLP:conf/lics/KobayashiLG19} shows that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
Note that $n$-PHORS is strictly less expressive then order-$n$ call-by-name PPCF:
% (because the underlying recursion schemes are not Turing complete):
for example, the PPCF terms in Examples \ref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
%\refExample{complexExample} 
are not definable as PHORS.

@@

Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems \ref{thm:antitone partial implies rankable,thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.

>>>>

\cite{DBLP:conf/lics/KobayashiLG19} shows that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
Note that $n$-PHORS is strictly less expressive then order-$n$ call-by-name PPCF:
% (because the underlying recursion schemes are not Turing complete):
for example, the PPCF terms in Examples \ref{ex:raven complex}, \ref{ex:non-affine continuous} and \ref{ex:higher-order recursion}
%\refExample{complexExample} 
are not definable as PHORS.

@@

Although this is a broadly applicable method of proving almost sure termination of probabilistic functional programs, and Theorems \ref{thm:antitone partial implies rankable} and \ref{thm:confluent ranking} make it more convenient to use, some method of automating the construction and checking of (antitone) ranking functions, even partially, could make it considerably more practically useful, especially in cases where almost sure termination is merely a side-condition for some other theorem or algorithm to be applicable.
